NanoMaLy

### 2015-11-17 paper/pldi16-submission/semantics.tex

\paragraph{Solution:} \rj{STOP}

@@

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{smallcheck}).
%
We track how the holes are incrementally filled in, by using
(value) \emph{substitutions} that map holes (denoted by their index)
to values which may, in turn, contain holes.
%

\subsection{Semantics}
\label{sec:semantics}
\input{operational.tex}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
and a substitution $\vsu$.
%
We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
%
We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
$e$ is $e_0$, $e'$ is $e_j$ and $\forall 0 \leq i < j$, we have
\step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
%
We write \steps{e}{\su}{e'}{\su'} if there exists some (finite) $j$ such that
$\stepi{j}{e}{\su}{e'}{\su'}$.

The evaluation relation is parameterized by a pair of \forcesym and
\gensym functions that handle runtime type-checking and instantiation of
holes respectively.
%
The relation must also maintain a substitution \su, mapping holes to
concrete values, so that we do not instantiate the same hole with
different values in different contexts, and so that we can report a
concrete witness to any discovered type errors.

Note that each primitive reduction step -- addition, if-elimination, and
function application -- uses \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation.
%
Additionally, beta-reduction \emph{does not} type-check its argument, it
only ensures that the value being applied is a function.
%
\begin{thm}
\label{thm:all-reduce}
  Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
  \ES{do we really need to state this, or is it obvious?}
\end{thm}
% \begin{proof}%[Proof of \autoref{thm:all-reduce}]
%   Simple induction on the evaluation relation.
% \end{proof}
%
\subsection{Soundness}
\label{sec:soundness}
We now show that our evaluation relation conservatively instantiates
holes, \ie if, given a function $f$, we report that
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$,
then for any type $\tfun{s}{t}$ you assign to $f$, there exists an input \hastype{v}{s} such that
$\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.

>>>>

\paragraph{Solution:} \RJ{STOP}

@@

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{smallcheck}).
%
We track how the holes are incrementally filled in, by using
(value) \emph{substitutions} that map holes (denoted by their index)
to values which may, in turn, contain holes.
%
The substitutions let us ensure that we instantiate each hole
with the same (partially defined) value, regardless of the multiple
contexts in which the hole appears, and hence, ensures we can
report a concrete (and general) witness for any (dynamically)
discovered type errors.

\subsection{Semantics}
\label{sec:semantics}
\input{operational.tex}
%
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{closed expression} is a pair $\pair{e}{\vsu}$ of an expression $e$
and a substitution $\vsu$.
%
We write $\step{e}{\vsu}{e'}{\vsu'}$ if the closed expression $\pair{e}{\vsu}$
transitions in a \emph{small step} to $\pair{e'}{\vsu'}$.
%
We write \stepi{j}{e}{\su}{e'}{\su'} if there exist $e_1,\ldots,e_j$ such that
$e$ is $e_0$, $e'$ is $e_j$ and $\forall 0 \leq i < j$, we have
\step{e_i}{\su_i}{e_{i+1}}{\su_{i+1}}.
%
We write \steps{e}{\su}{e'}{\su'} if there exists some (finite) $j$ such that
$\stepi{j}{e}{\su}{e'}{\su'}$.
%
The evaluation relation is parameterized by a pair of functions:
called \emph{narrow} (\forcesym) and \emph{generate} (\gensym),
that ``dynamically'' perform type-checking and hole-filling
respectively.

\paragraph{Narrowing Types} The function $\force{e}{t}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t}
by yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t},
or yields the stuck term if no such refinement is possible:
%
$$
\begin{array}{lcl}
\forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t} & \defeq & \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}} \\
\force{v}{\ehole{}}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

\paragraph{Generating Values} The function $\gen{t}$ takes as input a type
\texttt{t} and returns \RJ{HEREHERE}

$$
\begin{array}{lcll}
\gensym       & ::      & t \rightarrow v \\
\gen{\tint}   & \defeq  & n & \quad \text{non-deterministic} \\
\gen{\tbool}  & \defeq  & b & \quad \text{non-deterministic} \\
\gen{\tfun{t_1}{t_2}}   & \defeq & \efun{x}{\ehole{i}}, & \quad \text{$i$ is fresh} \\
\gen{\thole{}} & \defeq & \ehole{i}, & \quad \text{$i$ is fresh} \\
\end{array}
$$

%

Note that each primitive reduction step -- addition, if-elimination, and
function application -- uses \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation.
%
Additionally, beta-reduction \emph{does not} type-check its argument, it
only ensures that the value being applied is a function.
%
\begin{thm}
\label{thm:all-reduce}
  Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
  \ES{do we really need to state this, or is it obvious?}
\end{thm}

%
% \begin{proof}%[Proof of \autoref{thm:all-reduce}]
%   Simple induction on the evaluation relation.
% \end{proof}
%
\subsection{Soundness}
\label{sec:soundness}
We now show that our evaluation relation conservatively instantiates
holes, \ie if, given a function $f$, we report that
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$,
then for any type $\tfun{s}{t}$ you assign to $f$, there exists an input \hastype{v}{s} such that
$\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
